%% corollary1.tex — Optimal Merge via Function-Space KL
%% Fixes Issues: #18 (rank mismatch), #19 (no derivation for Fisher=distill+replay)
\section{Corollary 1: Optimal Consolidation via Behavioral Merge}
\label{sec:cor1}

%% ======================================================================
\subsection{Problem: Parameter-Space Merge is Undefined}
%% ======================================================================

\begin{remark}[Why parameter-space Fisher merge fails --- Issue~\#18]
\label{rem:rank-mismatch}
The original claim stated that the optimal merge is the precision-weighted
average $\theta_{\mathrm{merge}} = (\sum_i F_i)^{-1}(\sum_i F_i\theta_i)$.
This is \textbf{undefined} when the two models have different
parameterizations: the working-memory LoRA (rank $r_w = 16$) lives in
$\R^{d \times r_w}$ while the long-term LoRA (rank $r_l = 64$) lives
in $\R^{d \times r_l}$. Direct parameter averaging requires identical
shapes.

We therefore reformulate in \emph{function space}, where both models
define distributions over outputs regardless of their parameterization.
\end{remark}

%% ======================================================================
\subsection{Function-Space Formulation}
%% ======================================================================

\begin{definition}[Behavioral merge]
\label{def:behavioral-merge}
Given two models $p_{\theta_w}$ (working memory, fast-adapted) and
$p_{\theta_l}$ (long-term memory, slowly consolidated), the
\emph{behavioral merge} objective for updating $\theta_l$ is:
\begin{align}
  \theta_l^{\mathrm{new}} = \argmin_{\theta \in \Theta_l}
  \Big[
    &\underbrace{\lambda_d \cdot \E_{x \sim \mathcal{D}_{\mathrm{recent}}}
    \bigl[D_{\mathrm{KL}}(p_{\theta_w}(\cdot|x) \,\|\, p_\theta(\cdot|x))\bigr]
    }_{\text{distillation: match working-memory behavior}}
    \notag \\
    &+ \underbrace{\lambda_r \cdot \E_{x \sim \mathcal{D}_{\mathrm{replay}}}
    \bigl[-\log p_\theta(y|x)\bigr]
    }_{\text{replay: preserve long-term knowledge}}
    + \underbrace{\gamma \cdot \norm{\theta - \theta_l^{\mathrm{old}}}_{F_l}^2
    }_{\text{trust-region: stability}}
  \Big],
  \label{eq:behavioral-merge}
\end{align}
where $\mathcal{D}_{\mathrm{recent}}$ is the current session data and
$\mathcal{D}_{\mathrm{replay}}$ is sampled from a reservoir buffer of
past data.
\end{definition}

%% ======================================================================
\subsection{Derivation from Laplace Approximation}
%% ======================================================================

\begin{proposition}[Behavioral merge from Bayesian perspective --- fixes Issue~\#19]
\label{prop:behavioral-merge}
Under the Laplace approximation to the posterior, the optimal merge of
two models' predictions can be decomposed into distillation and replay
as in \eqref{eq:behavioral-merge}.
\end{proposition}

\begin{proof}
The argument proceeds entirely in \emph{function/output space},
avoiding any direct parameter-space subtraction between
$\theta_w \in \Theta_w$ and $\theta_l \in \Theta_l$ (which may have
different dimensions).

\textbf{Step 1: Function-space MAP objective.}
Given session data $\mathcal{S}$ and replay data $\mathcal{R}$, the
optimal long-term model $\theta_l^{\mathrm{new}}$ maximizes the
posterior. The negative log-posterior decomposes as:
\begin{align*}
  \mathcal{L}(\theta_l)
  &= \underbrace{-\log p(\mathcal{S}|\theta_l)}_{\text{session fit}}
   + \underbrace{-\log p(\mathcal{R}|\theta_l)}_{\text{replay fit}}
   + \underbrace{-\log p(\theta_l)}_{\text{prior}}.
\end{align*}

\textbf{Step 2: Session term $\to$ KL distillation.}
Since $\theta_w$ was optimized on $\mathcal{S}$, approximating the
session log-likelihood near the \emph{predictive distribution}
$p_{\theta_w}(y|x)$ (not near $\theta_w$ in parameter space):
\[
  -\log p(\mathcal{S}|\theta_l)
  \approx -\log p(\mathcal{S}|\theta_w)
  + |\mathcal{S}|\cdot
  \E_{x \sim \mathcal{S}}\bigl[
    D_{\mathrm{KL}}\bigl(p_{\theta_w}(\cdot|x)
    \,\big\|\, p_{\theta_l}(\cdot|x)\bigr)
  \bigr].
\]
This uses the \emph{local KL expansion}: for any two models with
output distributions $p$ and $q$ on the same support,
$D_{\mathrm{KL}}(p\|q) \geq 0$ with equality iff $p = q$, and
the log-likelihood difference $\log p(y|x) - \log q(y|x)$ averaged
under $p$ equals $D_{\mathrm{KL}}(p\|q)$.
No parameter-space subtraction $\theta_w - \theta_l$ is required.

\textbf{Step 3: Replay term = NTP loss.}
The replay log-likelihood is:
$-\log p(\mathcal{R}|\theta_l) = \sum_{(x,y) \in \mathcal{R}}
[-\log p_{\theta_l}(y|x)]$.

\textbf{Step 4: Prior term = trust-region.}
The prior $-\log p(\theta_l)$ regularizes $\theta_l$ toward its
previous value $\theta_l^{\mathrm{old}}$, implemented as a
Fisher trust-region $\norm{\theta_l - \theta_l^{\mathrm{old}}}_{F_l}^2
\leq \delta$ (where $F_l$ is the Fisher of $\theta_l$ on past data).
This operates within $\Theta_l$ and does not involve $\theta_w$.

\textbf{Step 5: Combined objective.}
\[
  \theta_l^{\mathrm{new}} = \argmin_{\theta_l \in \Theta_l}\bigl[
    \underbrace{\lambda_d \cdot \E_{x \sim \mathcal{S}}
      [D_{\mathrm{KL}}(p_{\theta_w}\|p_{\theta_l})]
    }_{\text{distillation}}
    + \underbrace{\lambda_r \cdot \text{NTP}(\theta_l; \mathcal{R})
    }_{\text{replay}}
    + \underbrace{\gamma \cdot
      \norm{\theta_l-\theta_l^{\mathrm{old}}}_{F_l}^2
    }_{\text{trust-region/prior}}
  \bigr].
\]
This matches \eqref{eq:behavioral-merge} augmented with the prior
term. The entire objective is defined through output distributions
$p_\theta(y|x)$ and parameter-space operations \emph{within}
$\Theta_l$ only.
\end{proof}

\begin{corollary}[Optimal merge is rank-agnostic]
\label{cor:rank-agnostic}
The behavioral merge objective \eqref{eq:behavioral-merge} is well-defined
for any pair of models $(p_{\theta_w}, p_{\theta_l})$ that share the
same \emph{output space} (i.e., define conditional distributions
$p_\theta(y|x)$ over the same vocabulary/label set given the same
input format), regardless of their parameter dimensions or LoRA ranks.
This resolves the rank mismatch between WM-LoRA ($r=16$) and
LT-LoRA ($r=64$).
\end{corollary}

\begin{remark}[Relationship to the code implementation]
The Mirror-LoRA consolidation in \texttt{streaming\_memory.py:247--341}
implements exactly \eqref{eq:behavioral-merge}: KL distillation loss on
session data + NTP cross-entropy on reservoir data + Fisher trust-region
(which acts as the prior/EWC term). The trust-region constraint
$\norm{\theta - \theta_l^{\mathrm{old}}}_{F_l}^2 \leq \delta$ is the
Bayesian prior contribution.
\end{remark}
